---
title: ''
author: "Carson James"
date: "3/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plotting
```{r}
# DNase data

# plot concentration as function of density:
plot(x = DNase$density, y = DNase$conc)
```

```{r}
# add axis labels
plot(x = DNase$conc, y = DNase$density,
xlab='DNase concentration (ng/ml)', ylab='Optical density')
```


```{r}
# change point symbol
plot(DNase$conc, DNase$density, pch=4,
xlab='DNase concentration (ng/ml)', ylab='Optical density')

```


# **03/07**

## Box Plots: 
An 'a'-th percentile of an **ordered** dataset is just the point in the dataset that has 'a'% of the data less than or equal to the point

Ex: data: 5.5, 5.56, 5.57, 5.78, 5.9, 6, 6.1, 6.15, 6.2, 6.3 
This data is ordered, 
- the 10th percentile is 5.5 since 10% of the data is less than or equal to 5.5
- the 50th percentile is 5.9 since 50% of the data is less than or equal to 5.9.
- 6.2 is the 90th percentile

The first, second and third quartiles of a data set are just the 25th, 50th and 75th percentiles of the data. So 25% of the data is less than or equal to the first quartile, ... etc

A box plot consists of two whiskers and a box. The edges of the box are the first and third quartiles, the whiskers extending from the edges of the box mark the min and max, and inside the box, the 2 quartile is marked.

## Fomulas:
In R you can denote a dependence between variables 
by using the '~' operator. 
If you have a variables, x and y and you are interested in y as a function of x, when plotting you can use the formula 'y ~ x'.
```{r}
#DNase

# make box plots for density for each run using a formula
# when the function takes formulas, you only need to give variable names, 
#i.e. var1 ~ var2 instead of data$var1, data$var2
# however, you need to supply the data to the function
boxplot(density ~ Run, data=DNase)

# to do boxplot with a single data set x do boxplot(x)
# outliers are denoted by circles

# generating data from non symmetric distribution
x <- rgamma(100, shape = .5) 
boxplot(x)


```

## histograms:

to make a histogram use the `hist` function
```{r}
hist(DNase$density)
```


# **03/09**

## ggplot2 package
ggplot2 lets us make pretty plots using layers.

```{r}
# install.packages("ggplot2")
library(ggplot2)
#plot DNase density ~ conc (denisty as function of conc) 

# recall plot from base plot function:
plot(x = DNase$conc, y = DNase$density)

# now with ggplot2
ggplot(data=DNase, mapping=aes(x=conc, y=density)) +
geom_point()
       


```



```{r}
# now boxplot

# base boxplot 
boxplot(density ~ Run, data=DNase)

# ggplot with boxplot layer
ggplot(DNase, aes(x=Run, y=density)) + geom_boxplot()
```


## faceting
faceting is breaking up data according to a particular variable.
```{r}
# base layer
ggplot(DNase, aes(x=conc, y=density)) +
# point layer
geom_point() + 
# facet layer
    facet_wrap(Run ~ ., ncol=4)

```

## chickwits data:

```{r}
head(chickwts)
#?chickwts
# feed is factor, weigth is numeric

# get levels
levels(chickwts$feed)

# make bar plot for feed levels
ggplot(chickwts, aes(x=feed)) + geom_bar()
```


## Fill by factor: 
```{r}
# fill argument to aes function:
# this assigns a color to each factor. 

# boxplot
ggplot(chickwts, aes(x=feed, y=weight, fill=feed)) +
geom_boxplot()

# dotplot 
ggplot(chickwts, aes(x=feed, y=weight, fill=feed)) +
geom_dotplot(binaxis='y', stackdir="center", dotsize=0.8)


```


## scatter with smoothing
```{r}
ggplot(DNase, aes(x=conc, y=density)) + geom_point()

ggplot(DNase, aes(x=conc, y=density)) + geom_point() + geom_smooth()
```

## Factor Review

```{r}
x <- c("a", "b")
is.factor(x)
x <- as.factor(x)
is.factor(x)
x
```

# **03/21**


## EM algorithm 
```{r}
#install.packages("mixtools")
library(mixtools) 

# generate data for simulation:

# initialize data vector of length 20000
n <- 20000
data <- rep(0, n)

# for each data sample, with prob .3, .1, .6 sample from a normal(-6, 2), normal(0,1) and normal(6,2) (second param is scale/sd)
set.seed(10)
for (i in (1:n)) {
    group <- sample(x = c(1, 2, 3), size = 1, prob = c(.3, .1, .6))
    if (group == 1) {
        data[i] <- rnorm(n = 1, mean = -6, sd = 2)
    } else if (group == 2) {
        data[i] <- rnorm(n = 1, mean = 0, sd = 1)
    } else {
        data[i] <- rnorm(n = 1, mean = 6, sd = 2)
    }
}

#vizualize data
hist(data, breaks = 50)

# em algorithm 
# k is the number of normal distributions in the mixture
em.mix = normalmixEM(x = data, k = 3)

#lambda is the mixing proportions
em.mix$lambda 

#means
em.mix$mu

# sigma is sd
em.mix$sigma
```


## Clustering

```{r}
# euclidean distance: distance between numeric vectors

# distance between points c(1, 2, -1) and c(0, 1, 3)
x <- c(1, 2, -1)
y <- c(0, 1, 3)

dist <- x-y
dist <- dist^2
dist <- sum(dist)
dist <- sqrt(dist)

# hamming distance: distance between binary vectors (entries are 0,1's)

# distance between c(1, 0, 0) and c(0, 1, 0)
x <- c(1, 0, 0)
y <- c(0, 1, 0)
dist <- sum(x != y)

```

# Kmeans:
```{r}
# kmeans:

#make data:
data <- matrix(c(5,6,1,  6,7,1,  6,5,1,   .5,0,2,  1,0,2, 0,.5,2), nrow = 6, ncol = 3, byrow = TRUE)

# make into data frame
data <- data.frame(data)
# change column names
names(data) <- c("length", "weight", "group")
# make group a factor type
data$group <- as.factor(data$group)

#plot 
library(ggplot2)
ggplot(data, aes(x = length, y = weight, color = group)) + geom_point()



# kmeans clustering:
set.seed(10)
data.cl <- kmeans(x = data, centers = 2)

# get labels
data.cl$cluster


# as numeric for arithmetic
labels.true <- as.numeric(data$group)
labels.est <- data.cl$cluster

# as factor for plotting
labels.true.fac <- as.factor(labels.true)
labels.est.fac <- as.factor(data.cl$cluster)

# plot

# true groups
ggplot(data, aes(x = length, y = weight, color = labels.true.fac)) + geom_point() 

# estimated groups
ggplot(data, aes(x = length, y = weight, color = labels.est.fac)) + geom_point()



#compare estimated clusters to actual clusters:
# install.packages("fossil")
library(fossil)

# good test question: 

#use rand.index function to compare labels:
# explanation: https://davetang.org/muse/2017/09/21/the-rand-index/
rand.index(c(1,1,1,  2,2), c(2,2,2, 3,3))
rand.index(c(1,1,1,  2,2, 3), c(2,2,3, 3,3, 4))

rand.index(labels.est, labels.true)
# rand.index == 1 is a perfect match

```

# iris dataset
```{r}
library(datasets)
data(iris)

# see what data looks like
head(iris)
```


```{r}

# fit clusters:
# remember to only include numeric data
set.seed(10)
iris.cl <- kmeans(x = iris[ , 1:4], centers = 3)

# estimate clusters (numeric)
cl.est <- iris.cl$cluster

# get true clusters (numeric)
cl.true <- as.numeric(iris$Species)

library(fossil)
rand.index(cl.true, cl.est)
# decent cluster result

cl.true
cl.est

# plot with same colors (must have same labels):

#looks like the labels were permuted. The estimated labels should be permuted via 2 -> 1, 3 -> 2 and 1 -> 3 (this is a guess assuming the estimates are good)
n <- length(cl.est)
for (i in (1:n)) {
    if (cl.est[i] == 2) {
        cl.est[i] <- 1
    } else if (cl.est[i] == 3) {
        cl.est[i] <- 2
    } else {
        cl.est[i] <- 3
    }
}

# get estimated clusters as factor
cl.est.fac <- as.factor(cl.est)

# get true clusters as factor
cl.true.fac <- as.factor(cl.true)

#plot x = sepal length and y = sepal width with color:
library(ggplot2)
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = cl.true.fac)) + geom_point()

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = cl.est.fac)) + geom_point()

```







```{r}

# attribute matrix:
animals <- matrix(c(0,1,1,1,1,  0,1,1,1,1,  0,1,0,1,1,  0,0,0,1,0,  1,0,0,1,1,  0,0,0,0,0), nrow = 6, ncol = 5, byrow = TRUE)

rnames <- c("human", "ape", "mouse", "snake", "hawk", "spider")
cnames <- c("feather", "tetrapod", "bipedal", "vertebrate", "endotherm") 

animals <- data.frame(animals, row.names = rnames)
names(animals) <- cnames

# get distance matrix:
# method = manhattan  gives hamming distance in this case
dist.animals = as.matrix(dist(animals, diag=TRUE, method="manhattan", upper=TRUE))
dist.animals

# see slides for next steps



# hclust function:

# distance object, method = "manhattan" is the hamming distance
dist.animals =  dist(animals, diag=TRUE, method="manhattan")

# hclust function, method = "mcquitty" is the weighted average 
clust.animals = hclust(dist.animals, method="mcquitty")
plot(clust.animals)

```
